A greedy algorithm for the cardinality-constrained college application problem

### 2022-08-23 paper_alma/body.tex
line edits

The static model considered in this study is arguable more hostile to students than the Pandora's Box problem. If, for example, an unlucky Pandora is rejected from a safety school at an early round of application, then she can compensate for the unexpected loss by pivoting to a more risk-averse application strategy. By contrast, the decisionmaker in our college application problem must commit at the outset to applying to every school in her portfolio. The admissions process used in the United States can be viewed as the concatenation of both problems: In the fall, students solve \eqref{headlineproblem} and send out a batch of applications. Then, upon observing their admissions outcomes in March, they use the Pandora strategy to pursue additional offers by applying to schools that offer rolling admissions.

@@

The college application problem possesses a somewhat unusual property that we call \emph{optimal risk elasticity:} Students with large budgets allocate a larger proportion of their portfolio to competitive reach schools than those without. This property implies the inoptimality of heuristic solutions that target a fixed distribution of reach and safety schools in the application portfolio. It also suggests a rational, rather than risk-averse, explanation for why low-income students are relatively unlikely to apply to reach schools. Next, we analyze the shadow cost of the budget constraint and profile the kinds of student who stand to benefit most from interventions (such as fee waivers) to eliminate barriers to application.

Throughout the following discussion, reach, target, and safety schools should be understood as relative concepts: A ``reach school'' is not necessarily an elite university, but rather a school at which a given student's admissions odds are lower than the other schools in her input data.

>>>>

The static model considered in this study is arguable more hostile to students than the Pandora's Box problem. If, for example, an unlucky Pandora is rejected from a safety school at an early round of application, then she can compensate for the unexpected loss by pivoting to a more risk-averse application strategy. By contrast, the decisionmaker in our college application problem must commit at the outset to applying to every school in her portfolio. The admissions process used in the United States can be viewed as the concatenation of both problems: In the fall, students solve \eqref{headlineproblem} and send out a batch of applications. Then, upon observing their admissions outcomes in March, they use the Pandora's Box strategy to pursue additional offers by applying to schools that offer rolling admissions.

@@

The college application problem possesses a somewhat unusual property that we call \emph{optimal risk elasticity:} Students with large budgets allocate a larger proportion of their portfolio to competitive reach schools than those without. This property implies the inoptimality of heuristic solutions that target a fixed distribution of reach and safety schools in the application portfolio. It also suggests a rational, rather than risk-averse, explanation for why low-income students are relatively unlikely to apply to reach schools. Next, we analyze the shadow cost of the budget constraint and profile the kind of student who stands to benefit most from an intervention, such as an application fee waiver, to eliminate barriers to application.

Throughout the following discussion, reach, target, and safety schools should be understood as relative concepts: A ``reach school'' is not necessarily an elite university, but rather a school with high utility and low admissions odds relative to the others in a \emph{given} student's input data.
